The Array of Long Baseline Antennas for Taking Radio Observations from the Sub-Antarctic

### 2020-10-29 albatros_paper.tex
HCC added new figure, updated comments

\author{H.~C.~Chiang$^{1,2,\dagger}$, T.~Dyson$^{1}$, E.~Egan$^{1}$,
  S.~Eyono$^{2}$, N.~Ghazi$^{2}$, J.~Hickish$^{3}$,
  J.~M.~J\'auregui--Garcia$^{4}$, V.~Manukha$^{5}$, T.~Moso$^{2}$,
  J.~Peterson$^{6}$, L.~Philip$^{7}$, J.~L.~Sievers$^{1,7}$,
  S.~Tartakovsky$^{1}$}

@@

To assess the loss of sensitivity as the signal level increases, we
use \autoref{eqn:1bit_output} to derive the correlator efficiency
$\epsilon$, which is the ratio of measured signal-to-noise ratio (SNR)
to the ideal, infinite-precision SNR.  \autoref{fig:1bit_efficiency}
illustrates that the decrease in $\epsilon$ as a function of SNR is
relatively shallow.  When the SNR is near zero, $\epsilon$ approaches
$2/\pi \approx 0.64$, and if the signal level increases to two times
the  
noise level, $\epsilon$ decreases by only a factor of $\sim2$.  The
final \albatros\ array will have baselines that are long relative to
the observing wavelength, and the correlated component of the electric
field (which sets the relevant signal level) is likely a small
fraction of the total received power.  This qualitative argument can
by made by considering antennas in the $uv$~plane in the flat-sky
approximation.  Any individual antenna receives all power present in
the $uv$~plane (plus system noise), but the correlated contribution
comes only from the portion of the $uv$~plane that is encompassed by
the primary beam at the antenna's $uv$-space coordinate.  Assuming
that the sky can be described as a Gaussian random field, the maximum
correlated signal between two antennas is determined by the aperture
filling factor of that pair.  For the case of dipole antennas, the
filling factor is the square of the wavelength divided by the baseline
length.  Therefore, in the long-baseline limit, the correlated signal
is small in comparison to the total power received, and the correlator
efficiency is unlikely to decrease significantly from the ideal
$2/\pi$ value.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/albatros2_waterfalls.jpg}
    \caption{Spectra from a co-aligned polarization pair from the
      two-element \albatros\ pathfinder, shown as a function of
      frequency and time.  The auto- and cross-spectra are shown in
      the top and bottom rows, respectively.  The spectrum magnitudes
      are in uncalibrated ADC units on a logarithmic scale, and the
      cross-spectrum is about two orders of magnitude fainter than the
      autospectra.  Repeatable structure from the Galaxy is visible
      over the three-day time scale in all plots.}
    \label{Fig:albatros2_waterfalls}
  \end{center}
\end{figure}

@@

The primary goal of the two-element pathfinder described in
\S\ref{s:2elem} was to qualitatively assess the level of RFI and
ionospheric contamination at low frequencies.
\autoref{Fig:albatros2_waterfalls} illustrates auto- and cross-spectra
from one of the co-aligned polarization pairs from the two antennas.
The data are shown over a three-day period (June 18--21, 2018), and no
calibration or cuts have been applied.  The rising and setting of the
Galaxy is visible in all of the spectrum magnitudes, and repeatable
fringes are visible in the phases of the cross-spectra.  The signal
drop-off below $\lesssim30$~MHz in the autospectra is caused largely
by the combined response of the LWA antenna and FEE.  Although the LWA
antenna is not optimized for the lowest \albatros\ observing
frequencies, as described in \S\ref{s:antenna}, it is sufficient for
these initial, exploratory measurements.  The RFI lines at
$\lesssim20$~MHz arise primarily from shortwave radio transmission
reflecting off the ionosphere.  During the night, when the ionospheric
plasma cutoff frequency drops, the maximum frequency of the RFI
contamination falls to $\sim10$~MHz.

@@

We gratefully acknowledge the National Research Foundation (grant
number 110929), the South African National Antarctic Programme, the
Natural Sciences and Engineering Research Council of Canada (grant
number RGPNS-2019-534549), and the Polar Continental Shelf Program for
providing funding and logistical support for our research program.
This research was undertaken, in part, thanks to funding from the
Canada 150 Program.  The financial assistance of the South
African SKA Project (SKA SA) towards this research is hereby
acknowledged (\url{www.ska.ac.za}).  We additionally extend our
sincere gratitude to the staff at SKA SA for hosting our lab tests.
We thank the South African National Space Agency for their technical
support, the crew of Ultimate Heli for safely delivering us and our
cargo, and all of the Marion takeover and winter team members who
provided invaluable advice and field help.  The authors also wish to
thank members of the LWA and RAPID collaborations for useful
discussions, and Raul Monsalve for helping make our first MARS
campaign a success.

>>>>

\author{H.~C.~Chiang$^{1,2,\dagger}$, T.~Dyson$^{1}$, E.~Egan$^{1}$,
  S.~Eyono$^{2}$, N.~Ghazi$^{2}$, J.~Hickish$^{3}$,
  J.~M.~J\'auregui--Garcia$^{4}$, V.~Manukha$^{5}$, T.~M\'enard$^{1}$,
  T.~Moso$^{2}$, J.~Peterson$^{6}$, L.~Philip$^{7}$,
  J.~L.~Sievers$^{1,7}$, S.~Tartakovsky$^{1}$}

@@

To assess the loss of sensitivity as the signal level increases, we
use \autoref{eqn:1bit_output} to derive the correlator efficiency
$\epsilon$, which is the ratio of measured signal-to-noise ratio (SNR)
to the ideal, infinite-precision SNR.  \autoref{fig:1bit_efficiency}
illustrates that the decrease in $\epsilon$ as a function of SNR is
relatively shallow.  When the SNR is near zero, $\epsilon$ approaches
$2/\pi \approx 0.64$, and if the signal level increases to two times
the noise level, $\epsilon$ decreases by only a factor of $\sim2$.
The final \albatros\ array will have baselines that are long relative
to the observing wavelength, and the correlated component of the
electric field (which sets the relevant signal level) is likely a
small fraction of the total received power.  This qualitative argument
can by made by considering antennas in the $uv$~plane in the flat-sky
approximation.  Any individual antenna receives all power present in
the $uv$~plane (plus system noise), but the correlated contribution
comes only from the portion of the $uv$~plane that is encompassed by
the primary beam at the antenna's $uv$-space coordinate.  Assuming
that the sky can be described as a Gaussian random field, the maximum
correlated signal between two antennas is determined by the aperture
filling factor of that pair.  For the case of dipole antennas, the
filling factor is the square of the wavelength divided by the baseline
length.  Therefore, in the long-baseline limit, the correlated signal
is small in comparison to the total power received, and the correlator
efficiency is unlikely to decrease significantly from the ideal
$2/\pi$ value.  This efficiency behavior remains true even when the
brightest radio sources are visible, since the primary beam
encompasses more than a steradian solid angle (the sun is irrelevant
since ionospheric effects restrict \albatros\ to night-time
observing).  Centaurus~A~\citep{culgoora1} and M1~\citep{Braude1969}
affect the total power expected from the Global Sky
Model~\citep{gsm2008} at only the $\sim$10\% level.  Additionally, as
will be shown in \S\ref{s:obs}, preliminary \albatros\ data
demonstrate that the visibilities are only a few percent of the total
power amplitude, thus indicating that diffuse emission dominates even
the brightest sources over the large primary beam.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/albatros2_waterfalls.jpg}
    \caption{Spectra from a co-aligned polarization pair from the
      two-element \albatros\ pathfinder, shown as a function of
      frequency and time.  The auto- and cross-spectra are shown in
      the top and bottom rows, respectively.  The spectrum magnitudes
      are in uncalibrated ADC units on a logarithmic scale, and the
      cross-spectrum amplitude is about two orders of magnitude
      fainter than the autospectra.  The cross-spectrum phase is shown
      in radians.  Repeatable structure from the Galaxy is visible
      over the three-day time scale in all plots.}
    \label{Fig:albatros2_waterfalls}
  \end{center}
\end{figure}

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.7\linewidth]{Figures/albatros2_auto_model.png}
    \caption{Top panel: simulated $S11$ for the LWA antenna,
      illustrating the steep signal loss below $\sim30$~MHz.  Bottom
      panel: median uncalibrated autospectrum for one of the
      polarizations in the two-element \albatros\ pathfinder, compared
      against a crude sky signal estimate given by the product of $(1
      - S11)$ with a nominal $\nu^{-2.6}$ synchrotron spectrum.  This
      simple model qualitatively demonstrates that the decrease in
      autospectrum power below $\sim30$~MHz is caused primarily by the
      antenna response.}
    \label{Fig:albatros2_auto_model}
  \end{center}
\end{figure}

@@

The primary goal of the two-element pathfinder described in
\S\ref{s:2elem} was to qualitatively assess the level of RFI and
ionospheric contamination at low frequencies.
\autoref{Fig:albatros2_waterfalls} illustrates auto- and cross-spectra
from one of the co-aligned polarization pairs from the two antennas.
The data are shown over a three-day period (June 18--21, 2018), and no
calibration or cuts have been applied.  The rising and setting of the
Galaxy is visible in all of the spectrum magnitudes, and repeatable
fringes are visible in the phases of the cross-spectra.  The signal
drop-off below $\lesssim30$~MHz in the autospectra is caused largely
by the combined response of the LWA antenna and FEE, as illustrated in
\autoref{Fig:albatros2_auto_model}.  Although the LWA antenna is not
optimized for the lowest \albatros\ observing frequencies, as
described in \S\ref{s:antenna}, it is sufficient for these initial,
exploratory measurements.  The RFI lines at $\lesssim20$~MHz arise
primarily from shortwave radio transmission reflecting off the
ionosphere.  During the night, when the ionospheric plasma cutoff
frequency drops, the maximum frequency of the RFI contamination falls
to $\sim10$~MHz.

@@

We gratefully acknowledge the National Research Foundation (grant
number 110929), the South African National Antarctic Programme, the
Natural Sciences and Engineering Research Council of Canada (grant
number RGPNS-2019-534549), and the Polar Continental Shelf Program for
providing funding and logistical support for our research program.
This research was undertaken, in part, thanks to funding from the
Canada 150 Program.  This research was enabled in part by support
provided by SciNet(https://www.scinethpc.ca/) and Compute Canada
(www.computecanada.ca).  The financial assistance of the South
African SKA Project (SKA SA) towards this research is hereby
acknowledged (\url{www.ska.ac.za}).  We additionally extend our
sincere gratitude to the staff at SKA SA for hosting our lab tests.
We thank the South African National Space Agency for their technical
support, the crew of Ultimate Heli for safely delivering us and our
cargo, and all of the Marion takeover and winter team members who
provided invaluable advice and field help.  The authors also wish to
thank members of the LWA and RAPID collaborations for useful
discussions, and Raul Monsalve for helping make our first MARS
campaign a success.
